---
title: "BCB 521 - A1: Literate Programming"
author: "Zoë Wilson"
format: html
editor: source
---

```{r}
#| include: false
library(tidyverse)
library(microbiome)
library(knitr)

data(atlas1006)

```

## Examination of Intestinal Microbiota of Western Adults.

### Summary

This script examines the relative abundance of phyla comprising the intestinal microbiota of adults living in the UK or USA.
Data used in this script is published in Lahti et al., 2014 and was obtained from and examined with the following packages:

```{r}
#| echo: false
paste0("microbiome ", packageVersion("microbiome"))

paste0("tidyverse ", packageVersion("tidyverse"))

paste0("knitr ", packageVersion("knitr"))

```

### Data Analysis

```{r Dat}
#| include: false

# Rename data
mdat <- atlas1006

# Save sample data to examine
sdat <- mdat@sam_data
```

#### Relative Abundance
Relative Abundance of the intestinal microbiota for 1006 individuals in the UK and/or USA varies by age, sex, nationality, and BMI.
```{r RA N}
#| echo: false
#| error: false
#| warning: false
mdat3 <- atlas1006 %>%
  subset_samples(!is.na(nationality)) %>%
  aggregate_taxa(level = "Phylum") %>%  
  microbiome::transform(transform = "compositional")

plot_composition(mdat3,
                      average_by = "nationality",
                      taxonomic.level = "Phylum",
                      sample.sort = "nationality",
                      x.label = "nationality") +
  scale_fill_brewer("Phyla", palette = "Paired") +
  guides(fill = guide_legend(ncol = 1)) +
  labs(x = "Nationality", y = "Relative abundance (%)",
       title = "Relative abundance data") + 
  theme(axis.text.x = element_text(angle=90, hjust=1),
        legend.text = element_text(face = "italic"))

```

```{r RA BMI}
#| echo: false
#| error: false
#| warning: false
mdat4 <- atlas1006 %>%
    subset_samples(!is.na(bmi_group)) %>%
  aggregate_taxa(level = "Phylum") %>%  
  microbiome::transform(transform = "compositional")

plot_composition(mdat4,
                      average_by = "bmi_group",
                      taxonomic.level = "Phylum",
                      sample.sort = "bmi_group",
                      x.label = "bmi_group") +
  scale_fill_brewer("Phyla", palette = "Paired") +
  guides(fill = guide_legend(ncol = 1)) +
  labs(x = "BMI", y = "Relative abundance (%)",
       title = "Relative abundance data") + 
  theme(axis.text.x = element_text(angle=90, hjust=1),
        legend.text = element_text(face = "italic"))

```

```{r RA S}
#| echo: false
#| error: false
#| warning: false
mdat5 <- atlas1006 %>%
  subset_samples(!is.na(sex)) %>%
  aggregate_taxa(level = "Phylum") %>%  
  microbiome::transform(transform = "compositional")

plot_composition(mdat5,
                      average_by = "sex",
                      taxonomic.level = "Phylum",
                      sample.sort = "sex",
                      x.label = "sex") +
  scale_fill_brewer("Phyla", palette = "Paired") +
  guides(fill = guide_legend(ncol = 1)) +
  labs(x = "Sex", y = "Relative abundance (%)",
       title = "Relative abundance data") + 
  theme(axis.text.x = element_text(angle=90, hjust=1),
        legend.text = element_text(face = "italic"))

```

```{r RA A}
#| echo: false
#| error: false
#| warning: false
mdat6 <- atlas1006 %>%
  subset_samples(!is.na(age)) %>%
  aggregate_taxa(level = "Phylum") %>%  
  microbiome::transform(transform = "compositional")

plot_composition(mdat6,
                      average_by = "age",
                      taxonomic.level = "Phylum",
                      sample.sort = "age",
                      x.label = "age") +
  scale_fill_brewer("Phyla", palette = "Paired") +
  guides(fill = guide_legend(ncol = 1)) +
  labs(x = "Age", y = "Relative abundance (%)",
       title = "Relative abundance data") + 
  theme(axis.text.x = element_text(angle=90, hjust=1),
        legend.text = element_text(face = "italic"))

```
